Computational methods in landscape ecology

### 2024-08-28 Manuscript.tex
adds my second affiliation
Update bib

  \author{Maximilian H.K. Hesselbarth}
            \affil{%
                  International Institute for Applied Systems Analysis,
                  Biodiversity, Ecology, and Conservation Group,
                  Laxenburg, Austria
              }
        \author{Jakub Nowosad}
            \affil{%
                  Adam Mickiewicz University, Poznan, Poland
              }
        \author{Marti Bosch}
            \affil{%
                  École polytechnique fédérale de Lausanne, Lausanne,
                  Switzerland
              }
        \author{Martin Jung}
            \affil{%
                  International Institute for Applied Systems Analysis,
                  Biodiversity, Ecology, and Conservation Group,
                  Laxenburg, Austria
              }
        \author{Rafael Schouten}
            \affil{%
                  Globe Institute, University of Copenhagen, Copenhagen,
                  Denmark
              }

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-Turner1989}
Turner MG (1989) Landscape ecology: The effect of pattern on process.
Annual Review of Ecology and Systematics 20:171--197.
\url{https://doi.org/10.1146/annurev.es.20.110189.001131}

@@

@@

@@

\bibitem[\citeproctext]{ref-Wu2002}
Wu J, Hobbs R (2002) Key issues and research priorities in landscape
ecology: An idiosyncratic synthesis. Landscape Ecology 17:355--365.
\url{https://doi.org/10.1023/A:1020561630963}

>>>>

  \author{Maximilian H.K. Hesselbarth}
            \affil{%
                  International Institute for Applied Systems Analysis,
                  Biodiversity, Ecology, and Conservation Group,
                  Laxenburg, Austria
              }
        \author{Jakub Nowosad}
            \affil{%
                  Adam Mickiewicz University, Poznan, Poland
              }
          \affil{%
                  University of Münster, Münster, Germany
              }
        \author{Marti Bosch}
            \affil{%
                  École polytechnique fédérale de Lausanne, Lausanne,
                  Switzerland
              }
        \author{Martin Jung}
            \affil{%
                  International Institute for Applied Systems Analysis,
                  Biodiversity, Ecology, and Conservation Group,
                  Laxenburg, Austria
              }
        \author{Rafael Schouten}
            \affil{%
                  Globe Institute, University of Copenhagen, Copenhagen,
                  Denmark
              }

@@

Despite the extensive use of the landscape mosaic model and the
landscape metrics based on it, major limitations have been identified
regarding both its ability to quantify the spatial structure of
landscapes and their relationship to ecological processes (Gustafson
1998; Li and Wu 2004; Dramstad 2009; Kupfer 2012; Lausch et al. 2015;
Frazier and Kedron 2017). In line with Dramstad (Dramstad 2009), we
review four major issues of landscape metrics, namely their sensitivity
to both spatial scale and thematic resolution, redundancy and
interpretation.

\subsubsection{Sensitivity to spatial scale and thematic
resolution}\label{sensitivity-to-spatial-scale-and-thematic-resolution}

Identifying the characteristic scale at which the organisms or processes
of interest operate is a critical first step towards establishing sound
relationships between pattern and process (Levin 1992; Wu 2007; Cushman
and McGarigal 2008). To that end, a key requirement it to understand the
sensitivity of landscape metrics to spatial scale in terms of both
resolution (grain) and extent.

A key concept to consider regarding grain size is the minimum mapping
unit (MMU), namely the smallest area entity that can be mapped and
analyzed as a discrete patch. Since the input maps should represent not
only the location of patches but also their shape, a grain size several
times smaller than the smallest patch (i.e., the MMU) in order to grasp
(even schematically) its shape. Even though the MMU is often assumed as
a natural characteristic of the dataset, it has an unsurprising impact
on the computed landscape metrics, especially in terms of the number of
patches and their size distribution (Saura 2002).

While the choice of grain is often constrained by the available data,
e.g., Landsat scenes at 30 m resolution, the selection of spatial extent
and the way that it is sampled varies largely depending on the
phenomenon of interest, e.g., from large transects of forests in natural
parks to small buffers along the urban-rural gradient. At the same time,
the choice of spatial extent and sampling strategy is often subject to
many arbitrary criteria that can largely influence the ultimate results.
Such an issue has long been recognized in physical geography as the
modifiable areal unit problem (MAUP) (Openshaw 1979, 1984). Approaches
to dealing with the MAUP in the context of landscape metrics include the
identification of individual entities that are ecologically meaningful
and not modifiable, the evaluation of ``optimal'' zoning systems that
maximize interzonal variations and minimize intrazonal variation and the
analysis of the sensitivity of the zoning configuration (Jelinski and Wu
1996; Duque et al. 2018; Fletcher Jr et al. 2023).

TODO: section about approaches to sample landscapes to compute metrics
and available software tools, e.g., zonal analysis with polygons (e.g.,
administrative limits), grids, transects or buffers along a landscape
gradient, moving windows (Bosch 2019; Hesselbarth et al. 2019; Huais
2024)?

The scale effects of landscape metrics can be assessed by computing
scalograms, i.e., the response curves of metrics to changing scales. For
instance, some metrics related to the number and density of patches and
edges show consistent linear and power-law scaling relationships to
changing grain across a wide diversity of simulated and real landscapes,
whereas other metrics exhibit less predictable responses (Turner et al.
1989; Saura and Martínez-Millán 2001; Wu et al. 2002; Wu 2004; Saura and
Castro 2007; Šı́mová and Gdulová 2012; Argañaraz and Entraigas 2014;
Frazier et al. 2021). Nonetheless, the effects of changing the sampling
extent can be less predictable than those of changing grain size (Wu
2004). For instance, in landscapes with small extent to grain ratio,
boundary effects are likely to dominate and truncate patches, resulting
in unreliable landscape metric computations. In order to address scale
issues and advance the science of scaling in landscape ecology, studies
should explicitly consider the concept of scope, namely the ratio of
extent to resolution (Schneider 2001). As suggested by Frazier (Frazier
2023), this can be done following three key steps, namely by considering
scope guidelines when designing experiments, explicitly reporting grain
and extent so that scope can be calculated allowing quantitative
comparability with other studes and undertaking comparisons and
replications based on scope. The latter is showcased in a proof of
concept example experiment, where landscapes from the same ecoregion are
sampled using different grain and extent and then the distributions of
the computed landscape metrics are visualized through ridgeline plots
colored by scope (Frazier 2023).

TODO: section about landscape metric rescaling (Saura and Castro 2007;
Argañaraz and Entraigas 2014; Frazier 2014, 2016; Frazier et al. 2021)?

TODO: mention hierarchichal patch dynamics (Wu 1999)?

TODO: sensitivity to thematic resolution (Huang et al. 2006; Bailey et
al. 2007; Buyantuyev and Wu 2007; Castilla et al. 2009).

\subsubsection{Redundancy}\label{redundancy}

TODO

\begin{itemize}
\item
  Factor analysis of Riitters et al. (Riitters et al. 1995) found six
  main components: average patch compaction, image texture, average
  patch shape, patch perimeter-area scaling, number of attribute classes
  and large-patch density-area scaling.
\item
  Cushman et al. (Cushman et al. 2008) not only isolate the components
  but also assess their strength (proportion of variance explained),
  universality (how present they are accross landscapes) and consistency
  (how stable they are accross landscapes). At the class level, they
  find seven ``highly universal and consistent structure components''
  (page 700), namely edge contrast, patch shape complexity, aggregation,
  nearest neighbor distance, patch dispersion, large patch dominance and
  neighborhood similarity. At the landscape level, five components are
  identified, namely contagion/diversity (similar to image texture in
  Riitters et al. (Riitters et al. 1995), large patch dominance,
  interspersion/juxtaposition, edge contrast, patch shape variability,
  proximity and nearest neighbor distance. Overall, components are more
  consistent and universal at the landscape level than at the class
  level.
\item
  Uuemaa et al. (Uuemaa et al. 2011) find four main components:
  dominance (also identified by Riitters et al. (Riitters et al. 1995)
  and Cushman et al. (Cushman et al. 2008)), contrast, shape complexity
  and composition.
\item
  Amount (composition) and adjacency (configuration) are the fundamental
  elements of pattern (Riitters 2019)
\item
  Entropy (Nowosad and Stepinski 2019)
\item
  Spatial signatures (Nowosad 2021)
\end{itemize}

\subsubsection{Interpretability}\label{interpretability}

TODO - do landscape metrics measure fragmentation? - normalized metrics
(Wang and Cumming 2011) - (Frazier et al. 2021)

Responses: - incorporating aspects of ecological function into landscape
metrics (e.g., core area, least-cost distances) and moving towards
graph-based and circuit theory approaches (Kupfer 2012). - point-data
analysis (Gustafson 1998) and gradient surface model (Lausch et al.
2015) require fewer assumptions regarding LULC categorization schemes or
spatial scale

@@

\bibitem[\citeproctext]{ref-Arganaraz2014}
Argañaraz JP, Entraigas I (2014) Scaling functions evaluation for
estimation of landscape metrics at higher resolutions. Ecological
informatics 22:1--12

\bibitem[\citeproctext]{ref-Bailey2007}
Bailey D, Billeter R, Aviron S, et al (2007) The influence of thematic
resolution on metric selection for biodiversity monitoring in
agricultural landscapes. Landscape Ecology 22:461--473

@@

\bibitem[\citeproctext]{ref-Bosch2019}
Bosch M (2019) PyLandStats: An open-source pythonic library to compute
landscape metrics. PLoS One 14:e0225734

\bibitem[\citeproctext]{ref-Buyantuyev2007}
Buyantuyev A, Wu J (2007) Effects of thematic resolution on landscape
pattern analysis. Landscape Ecology 22:7--13

@@

\bibitem[\citeproctext]{ref-Castilla2009}
Castilla G, Larkin K, Linke J, Hay GJ (2009) The impact of thematic
resolution on the patch-mosaic model of natural landscapes. Landscape
Ecology 24:15--23

@@

\bibitem[\citeproctext]{ref-Cushman2008Landscape}
Cushman SA, McGarigal K (2008) Landscape metrics, scales of resolution.
In: Designing green landscapes. Springer, pp 33--51

\bibitem[\citeproctext]{ref-Cushman2008Parsimony}
Cushman SA, McGarigal K, Neel MC (2008) Parsimony in landscape metrics:
Strength, universality, and consistency. Ecological indicators
8:691--703

@@

\bibitem[\citeproctext]{ref-Dramstad2009}
Dramstad WE (2009) Spatial metrics--useful indicators for society or
mainly fun tools for landscape ecologists? Norsk Geografisk
Tidsskrift-Norwegian Journal of Geography 63:246--254

@@

\bibitem[\citeproctext]{ref-Duque2018}
Duque JC, Laniado H, Polo A (2018) S-maup: Statistical test to measure
the sensitivity to the modifiable areal unit problem. PloS one
13:e0207377

@@

\bibitem[\citeproctext]{ref-Fletcher2023}
Fletcher Jr RJ, Betts MG, Damschen EI, et al (2023) Addressing the
problem of scale that emerges with habitat fragmentation. Global Ecology
and Biogeography 32:828--841

@@

\bibitem[\citeproctext]{ref-Frazier2014}
Frazier AE (2014) A new data aggregation technique to improve landscape
metric downscaling. Landscape ecology 29:1261--1276

\bibitem[\citeproctext]{ref-Frazier2016}
Frazier AE (2016) Surface metrics: Scaling relationships and downscaling
behavior. Landscape ecology 31:351--363

\bibitem[\citeproctext]{ref-Frazier2023}
Frazier AE (2023) Scope and its role in advancing a science of scaling
in landscape ecology. Landscape Ecology 38:637--643

@@

\bibitem[\citeproctext]{ref-Frazier2021}
Frazier AE, Kedron P, Ovando-Montejo GA, Zhao Y (2021) Scaling spatial
pattern metrics: Impacts of composition and configuration on downscaling
accuracy. Landscape Ecology 1--16

@@

\bibitem[\citeproctext]{ref-Gustafson1998}
Gustafson EJ (1998) Quantifying landscape spatial pattern: What is the
state of the art? Ecosystems 1:143--156

@@

\bibitem[\citeproctext]{ref-Huais2024}
Huais PY (2024) Multilandr: An r package for multi-scale landscape
analysis. Landscape Ecology 39:140

\bibitem[\citeproctext]{ref-Huang2006}
Huang C, Geiger E, Kupfer J (2006) Sensitivity of landscape metrics to
classification scheme. International Journal of Remote Sensing
27:2927--2948

@@

\bibitem[\citeproctext]{ref-Jelinski1996}
Jelinski DE, Wu J (1996) The modifiable areal unit problem and
implications for landscape ecology. Landscape ecology 11:129--140

@@

\bibitem[\citeproctext]{ref-Kupfer2012}
Kupfer JA (2012) Landscape ecology and biogeography: Rethinking
landscape metrics in a post-FRAGSTATS landscape. Progress in physical
geography 36:400--420

@@

\bibitem[\citeproctext]{ref-Levin1992}
Levin SA (1992) The problem of pattern and scale in ecology: The robert
h. MacArthur award lecture. Ecology 73:1943--1967

\bibitem[\citeproctext]{ref-Li2004Use}
Li H, Wu J (2004) Use and misuse of landscape indices. Landscape ecology
19:389--399

@@

\bibitem[\citeproctext]{ref-Openshaw1979}
Openshaw S (1979) A million or so correlated coefficients: Three
experiment on the modifiable areal unit problem. Statistical
applications in the spatial sciences

\bibitem[\citeproctext]{ref-Openshaw1984}
Openshaw S (1984) Ecological fallacies and the analysis of areal census
data. Environment and planning A 16:17--31

@@

\bibitem[\citeproctext]{ref-Riitters2019}
Riitters K (2019) Pattern metrics for a transdisciplinary landscape
ecology. Landscape Ecology 34:2057--2063

\bibitem[\citeproctext]{ref-Riitters1995}
Riitters KH, O'neill R, Hunsaker C, et al (1995) A factor analysis of
landscape pattern and structure metrics. Landscape ecology 10:23--39

@@

\bibitem[\citeproctext]{ref-Saura2002}
Saura S (2002) Effects of minimum mapping unit on land cover data
spatial configuration and composition. International Journal of Remote
Sensing 23:4853--4880

\bibitem[\citeproctext]{ref-Saura2007}
Saura S, Castro S (2007) Scaling functions for landscape pattern metrics
derived from remotely sensed data: Are their subpixel estimates really
accurate? ISPRS Journal of Photogrammetry and Remote sensing 62:201--216

\bibitem[\citeproctext]{ref-Saura2001}
Saura S, Martínez-Millán J (2001) Sensitivity of landscape pattern
metrics to map spatial extent. Photogrammetric Engineering \& Remote
Sensing 67:1027--1036

\bibitem[\citeproctext]{ref-Schneider2001}
Schneider DC (2001) The rise of the concept of scale in ecology: The
concept of scale is evolving from verbal expression to quantitative
expression. BioScience 51:545--553

@@

\bibitem[\citeproctext]{ref-Simova2012}
Šı́mová P, Gdulová K (2012) Landscape indices behavior: A review of scale
effects. Applied geography 34:385--394

@@

\bibitem[\citeproctext]{ref-Turner1989Landscape}
Turner MG (1989) Landscape ecology: The effect of pattern on process.
Annual Review of Ecology and Systematics 20:171--197.
\url{https://doi.org/10.1146/annurev.es.20.110189.001131}

@@

\bibitem[\citeproctext]{ref-Turner1989Effects}
Turner MG, O'Neill RV, Gardner RH, Milne BT (1989) Effects of changing
spatial scale on the analysis of landscape pattern. Landscape ecology
3:153--162

\bibitem[\citeproctext]{ref-Uuemaa2011}
Uuemaa E, Roosaare J, Oja T, Mander Ü (2011) Analysing the spatial
structure of the estonian landscapes: Which landscape metrics are the
most suitable for comparing different landscapes? Estonian Journal of
Ecology 60:70

@@

\bibitem[\citeproctext]{ref-Wang2011}
Wang X, Cumming SG (2011) Measuring landscape configuration with
normalized metrics. Landscape Ecology 26:723--736

@@

\bibitem[\citeproctext]{ref-Wu1999}
Wu J (1999) Hierarchy and scaling: Extrapolating information along a
scaling ladder. Canadian journal of remote sensing 25:367--380

\bibitem[\citeproctext]{ref-Wu2004}
Wu J (2004) Effects of changing scale on landscape pattern analysis:
Scaling relations. Landscape ecology 19:125--138

\bibitem[\citeproctext]{ref-Wu2007}
Wu J (2007) Scale and scaling: A cross-disciplinary perspective. In: Key
topics in landscape ecology. Cambridge University Press, pp 115--142

\bibitem[\citeproctext]{ref-Wu2002Key}
Wu J, Hobbs R (2002) Key issues and research priorities in landscape
ecology: An idiosyncratic synthesis. Landscape Ecology 17:355--365.
\url{https://doi.org/10.1023/A:1020561630963}

\bibitem[\citeproctext]{ref-Wu2002Empirical}
Wu J, Shen W, Sun W, Tueller PT (2002) Empirical patterns of the effects
of changing scale on landscape metrics. Landscape ecology 17:761--782

### 2024-08-28 Sections/01-introduction.qmd
Update bib

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].

>>>>

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989Landscape; @Turner2005].
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002Key; @Wu2013].
